Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-10-10 ms/review-responses.tex

\begin{point}{Line 35:}
    Also cite Wilkins and Wakeley, Genetics 2002; Wilkins 2004
\end{point}

\reply{
Done.
}

\begin{point}{\revref}
    "Such models have been used extensively in ecological modeling but rarely in population genetics " Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
\end{point}

\cjb{Peter - did you have any in mind here? I can come up with a couple but don't know the ecology lit on this very well}
\reply{

}

@@

\begin{point}{\revref}
    Please describe computation time needed per replicate
\end{point}

\reply{
We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
}

@@

\begin{point}{Figure 4:}
    Show random-mating expectation
\end{point}

\reply{
    \cjb{working on this}
}

\begin{point}{Figure 3A, S2:}
    Perhaps more revealing to show on log-log scale?
\end{point}

\reply{
Good suggestion -- the SFS in Fig 3A is now on a log-log scale, which shows the slight decrease in low frequency SNPs a little better. 
}

@@

\begin{point}{Figure S3:}
    Caption seems to be missing detail 
\end{point}

\reply{
    Thanks for catching this - we have revised this caption to add details including the simulation parameters. \llname{ll:figs3caption}
}

@@

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
The mean of the IBS tract distribution is calculated by building a (very large) vector of lengths of all IBS tracts between all pairs of individuals and taking the mean over this entire vector. This distribution is quite strongly skewed (i.e. see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values. We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts. The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts. Perhaps confusing we also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals. 
\cjb{need to think of something to say we did about this in the text}

}

@@

\begin{point}{}
    The authors use K to denote both carrying capacity \llname{ll:K_defn}
    and population density \llname{ll:K_again}. 
    It might be better to use a different notation for these quantities since carrying capacity is fixed while density is an emergent quantity in the non-Wright-Fisher model. Use of K to denote carrying capacity and density is a bit confusing. For example, on \llname{ll:K_more} it is said that ' the "population density" (K) and "mean lifetime" (L) parameters were the same in all simulations'. Here K seems to indicate carrying capacity rather than density? The latter is an emergent quantity and varies across simulation runs?
\end{point}

\reply{
    We agree that this distinction is worth emphasizing!
    We've adjusted our language to hopefully remind the reader that
    $K$ is a parameter that controls population density, rather than being equal to it,
    at \llname{ll:K_again} and \revref{} and \llname{ll:K_more}.
}

\begin{point}{}
    Concerning the non-Wright-Fisher model used, it would be helpful to emphasize that some of the parameters are emergent in contrast to Wright-Fisher model. For example, on Page 11, lines 306-308, the author's goal was to look at census size variation and variation in other quantities. This would be better understood if to emphasize that these parameters are emergent properties in the non-Wright-Fisher model used.
\end{point}

\reply{
    We have added to the text at the beginning of the results to emphasize that this analysis is necessary because these parameters are emergent rather than fixed \llname{emergentparams}. 
}

@@

\begin{point}{}
    Page 9, line 242, Perhaps 'Demographic Inference' might better reflect the content of this section.
\end{point}

\reply{
Good suggestion -- we have changed the section heading to "demographic inference" 
}

@@

\begin{point}{\revref}
    This sentence with 'Gaussian noise with mean zero and standard deviation 10' is confusing since it was mentioned earlier that the modelled phenotype must vary as human height across Europe, and human height varies 2 standard deviations. Only after reading the whole paragraph it becomes clear that 'standard deviation 10' here referes to unit of height. Please consider rephrasing this sentence.
\end{point}

\reply{
We have revised this sentence to clarify that we aim to produce a variation in mean phenotype of two standard deviations across the landscape \llname{phenotypescalingsd}.
}

@@

\begin{point}{\revref}
    In the sentence, 'We also examined p values for systemic inflation' I think the authors meant 'systematic inflation'.
\end{point}

\reply{
    Whoops; thanks. Fixed.
}

\begin{point}{}
    Page 11, Please correct the legend in Figure 2: must be 'spatial model' and 'random mating' model.
\end{point}

\reply{
ah, a confusing legend title placement! we moved "model" to after "spatial" as suggested.
}

@@

\begin{point}{}
    Optional: a dashed line in Figure 2 that shows the total carrying capacity of 50*50*5=12500 would be helpful.
\end{point}

\reply{
    This is a good suggestion, but we decided to not include this as we don't have straightforward expectations for the other parameters shown. 
}

@@

\begin{point}{}
    Please add or correct references to supplementary figures: For example, Figure S2 was probably meant to accompany Figure 3A, while Figure S1 Figure 3B, but references in the text are absent. In fact, the first reference is made to Figure S3 on page 15.
\end{point}

\reply{
\cjb{check on final pass}
}

\begin{point}{}
    There are also several typos and errors in the text. For example, on Page 12, lines 309; Page 27, line 655.
\end{point}

\reply{
    Thank you for noting these -- they have been corrected. 
}

@@

    \begin{enumerate}
        \item It nice to actually see spatially explicit simulations and I'm happy that forward simulation is now fast enough that you can do this sort of thing.
        \item The paper is very clear and well-written, easy to understand the motivation and most of the details. That's not always the case for this sort of paper.
        \item I felt that the section about the effect on GWAS was the most interesting and novel part of the paper and gave me some intuition that I hadn't had before.
    \end{enumerate}

    I don't have any major criticisms. There were a few aspects that I thought might warrant some additional discussion, and a few specific questions below. The general questions I had after reading it were:
\end{quote}

\begin{point}{}
    To what extent are any of the results dependent on the exact method of simulation. There are a number of choices about the exact details of the simulations (e.g. the way the overlapping generations are handled, the edge effects and, particularly, the form of Equation 1 - see below). It's not so much that these are non-standard (since I don't think there is a standard) and they all sort of make sense heuristically, and I was left wondering whether these sorts of choices actually make a difference. Do the authors have some thoughts/intuition/results about that? Given that the results in Fig. 3 seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like Fig. 2 which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
\end{point}

\reply{
This is an excellent point also raised by other reviewers. We have added an appendix section comparing our model to a reverse-time stepping stone analysis on several relevant summary statistics \llname{appendix1}. Briefly, we find that many of the qualitative trends we see also occur in a stepping-stone setting, suggesting that they are driven purely by the particularities of our parameterization. We also see several interesting departures from the stepping-stone model reflecting shortcomings in the gridded approach to approximating continuous landscapes, which we think our model helps to avoid. We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work. 
}

@@

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    See above. It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$ (at least in reverse-time simulations). 
}

@@

\begin{point}{}
    The source of equation (1) is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
    \cjb{peter - any suggestions here?}
}

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does "roughly" mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach. MAKE THIS FIG AND POINT TO IT HERE.
}

@@

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach. As you can see in \ref{fig:slim_v_msp_sfs}, these approaches yield extremely similar spectra so we believe any error in the approximation is quite small. We also now show variation in simulation times with different mutation and tree sequence recording schemes in Figure \ref{fig:runtimes}, and included a brief discussion of this in \llname{runtimesxn}
}

@@

\reply{

}

\begin{point}{}
    Fig. 7D: I am surprised by the extent to which the observed values of -log10(p) fall below the y=x line. Particularly in the lower right panel for large neighbourhood sizes. I would expect that to be close to panmictic - why are the P-values underdispersed? That seems like a potential bug, or else something weird is going on.
\end{point}

\reply{
    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion. It seems to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype. \cjb{Peter or Andy -- any advice here? } 
}

@@

\begin{point}{}
    Lines 706-716, It might be worth citing Haworth et al Nature Communications 2019 (https://doi.org/10.1038/s41467-018-08219-1) who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
\end{point}

\reply{
Done - thank you for pointing us to this study. \llname{haworthcite}
}

@@

\begin{point}{}
    The analysis and discussion around the effect of GWAS is focused on PCA correction. Do mixed models help at all?
\end{point}

\reply{
    We are also very interested to know how mixed models perform here, but think that adding a second GWAS method would make this section too large for the current paper. We have added a note to \llname{mixedmodelcallout} specifically citing mixed models as alternate methods that may perform better. 
}

@@

\begin{point}{}
    The github link to the code didn't work for me. I assume it will be made public later, but at this point I can't tell whether the code is available/useable.
\end{point}

\reply{
Apologies, I had forgotten to make the repo public. The link should work for everyone now. 
}

@@

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's "pain in the torus' article, 1975) or, more recently, the
    ``mugration' or ``discrete trait analysis' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

\begin{point}{}
    First, the simulation starts with individuals distributed uniformly at random in space. Is there any indication that the three-step algorithm used here maintains this distribution during the course of evolution? If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it? I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping' of individuals that hampers the Wright-Malecot model. Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
\end{point}

\reply{
    This is a good point. We evaluated this while developing the simulation using the built-in visualization tools of the SLiM GUI, and have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance. 
}

@@

Second, the demographic process used here involves birth and death of individuals. Does the population survive asymptotically or, like any birth-death process, eventually dies with probability one? In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
\begin{point}{}

\end{point}

\reply{
    Similarly here we monitored the population size over time while running SLiM. All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed. \cjb{is there something else we should show here?}
}

@@

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
    \cjb{Peter -- ideas here?}
}

\begin{point}{}
    Also, the web page https://github.com/petrelharp/spaceness does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
\end{point}

\reply{
I apologize -- I had forgotten to set the repo to public. It should work now. 
}

@@

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (https://github.com/jeromekelleher/discsim)
\end{point}

\reply{
    This is a good point also raised by other reviewers. We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models. See Appendix 1. We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

@@

\begin{point}{}
    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
    Census population in the random mating model scales with "neighborhood size" because of two effects -- the edge scaling and the competition

    \cjb{need a better answer here...}
}

@@

\begin{point}{}
    Line 729-731: "Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]". Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations? It would also probably be useful to mention that neighborhood sizes given in Table 1 should be compared with extreme caution since the size of the corresponding habitats vary across species. More generally, I suspect that the size of the habitat has a substantial impact on the vast majority of statistics examined in this study. Indeed, the mean parent-offspring distance, which is at the core of the definition of Wright's neighborhood size, is only small or large relative to the size of the habitat.
\end{point}

\reply{
This is a good point. Wright's work \citep{Wright1943} suggests some aspects of genetic variation such as variance in allele frequencies and inbreeding coefficients can be estimated by looking only at what he would later \citep{Wright1946} call "neighborhood size", but certainly other aspects like the number of segregating sits will also depend on total landscape size. We now note on \llname{popsizecaveat} that we have evaluated only one landscape size, and have added a sentence to the discussion noting that exploration of these patterns in varying landscape sizes is an important avenue for further research \llname{landscapesizecaveat}.

}

@@

%%%%%%%%%%%%%%%%%%%
\reviewersection{4}

\begin{quote}
    The manuscript by Battey et al explores the consequence of a well-known
    violation to population genetic models: the fact that populations are
    spatially structured and mate along a geographical cline, rather than
    randomly. This topic is important, particularly in light of recent working
    describing how spatially correlated genetic and environmental impacts can
    confound some population genetic insights, such as positive selection for
    height in Europe. The analyses and investigations presented here are
    thorough and sensible, and my comments are primarily intended to broaden
    accessibility for this interesting topic.
\end{quote}

\begin{point}{}
    Introduction. The discussion is very clear, articulating the three primary goals of the project: the impact of failing to model spatial population structure on 1) population genetic summary statistics, 2) inference on demographic history from population genetic data, and 3) impacts on GWAS summary statistics. I found the discussion a bit easier to follow than the introduction and would suggest streamlining and introducing the topic a bit more. Since the paper follows the flow described in the discussion, it might help orient readers by introducing these topics in the same order.
\end{point}

\reply{
    \cjb{working on this}
    Thank you for this suggestion. We have slightly revised the introduction and hope it is now clearer; however since we want to cover a little history and motivation for our continuous model vs stepping-stone approaches in the intro it does have a different flow from the discussion.

}

### 2019-10-10 ms/spaceness.tex
edits to andy and peter

\documentclass[10pt,twoside,lineno,hidelinks]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}. Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure XXXXX. Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
\label{fig:samplemap}
\end{figure}

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10 \revpoint{1}{5}
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

%%%%%%%%%%%
\subsection{Demographic Parameters}

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number.
These are shown in Figure~\ref{fig:genparams}.
This occurs because,
even through the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
were the same in all simulations,
the strength of stochastic effects depends strongly on $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

@@

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
\caption{Site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=.5\textwidth]{figures/ibs_dists2.pdf}
\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes ($N_{W}$).}
\label{fig:ibs}
\end{figure*}

@@

%%%%%%%%%%%
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes.}
\label{fig:dafdists}
\end{figure*}

@@

We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}). This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

@@

\section{Supplementary Figures and Tables}
\beginsupplement

>>>>

\documentclass[10pt,twoside,lineno,hidelinks]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1}
\usepackage{placeins}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}. Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs}. Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

To check that our model produces reasonable results, we compared its output to that of a stepping-stone model implemented in msprime \citep{Kelleher2016}. These results are discussed in detail in Appendix 1, but in general we find that our model produces many of patterns generated by stepping stone models while avoiding some artifacts associated with discretization of the landscape.

@@

\begin{figure}[htbp]
\centering
\includegraphics{ms/figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
\label{fig:samplemap}
\end{figure}

@@

%%%%%%%%%%%
\subsection{Demographic Inference}

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

%%%%%%%%%%%
\subsection{Demographic Parameters and Run Times}

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}. Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set \llabel{emergentparams}.
Variation across runs occurs because,
even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
were the same in all simulations,
the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/runtimes.pdf}
\caption{Run times of continuous space simulations with landscape width 50 and expected density 5 under varying neighborhood size. Times are shown for simulations run with mutations applied directly in SLiM (dashed lines) or later applied to tree sequences with msprime (solid lines). Times for simulations run with tree sequence recording disabled are shown in grey.}
\label{fig:runtimes}
\end{figure}

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to two weeks of computation. At small neighborhood sizes the overhead caused by processing mutations and simplifying tree sequences can add $\approx30\%$ to the total run time, but above neighborhood sizes of $\approx$ 100 the effects of neighborhood size dominate because our simulation requires pairwise calculations for sets of individuals within a $3\sigma$ radius. As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales is possible. \llabel{runtimesxn}

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics \llabel{spatialgenetic}.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{ms/figures/sfs_w_sumstats_loglog.pdf}
\caption{Log-scaled site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=.5\textwidth]{ms/figures/ibs_cumdists_w_rm.pdf}
\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).}
\label{fig:ibs}
\end{figure*}

@@

%%%%%%%%%%%
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes (NS).}
\label{fig:dafdists}
\end{figure*}

@@

We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. \llabel{ibdnesxn} In general methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events, but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short. Combined with our analysis of IBS tract length variation (Figure \ref{fig:ibs}) this suggests that haplotype-based methods are likely to be affected by similar biases.

A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}). This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

@@

\beginsupplement
\section{Supplementary Figures and Tables}
\FloatBarrier

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{ms/figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:slim_v_msp_sfs} 
\end{figure*}
\FloatBarrier
